---
title: "ch7-ex-applied"
author: "S Zimine"
date: "12/5/2017"
output: html_document
---


### Ex 6
#### 6.a
From Asad solution

```{r}
library(ISLR)
library(boot)

dfmax <- 10
kcv <- 10
all.deltas <- rep(NA, dfmax)
for (i in 1:dfmax) {
  glm.fit <- glm(wage~poly(age, i,raw=TRUE), data=Wage)
  all.deltas[i] = cv.glm(Wage, glm.fit, K=kcv)$delta[2]
}

plot(1:dfmax, all.deltas, xlab="Degree", ylab="CV error", type="l", pch=20, lwd=2, main="test MSE as a function of polynomial degree in wage~poly(wage,i)")
min.point = min(all.deltas)
sd.points = sd(all.deltas)
abline(h=min.point + 0.2 * sd.points, col="red", lty="dashed")
abline(h=min.point - 0.2 * sd.points, col="red", lty="dashed")
legend("topright", "0.2-standard deviation lines", lty="dashed", col="red")
```

The plot indicates that $df=3$ is a minimal degrees of freedom with a small enough test MSE.

Lets compare it with `anova()` results.

```{r}
fit1 <- lm(wage~poly(age,1,raw=TRUE),data=Wage)
fit2 <- lm(wage~poly(age,2,raw=TRUE),data=Wage)
fit3 <- lm(wage~poly(age,3,raw=TRUE),data=Wage)
fit4 <- lm(wage~poly(age,4,raw=TRUE),data=Wage)
fit5 <- lm(wage~poly(age,5,raw=TRUE),data=Wage)
fit6 <- lm(wage~poly(age,6,raw=TRUE),data=Wage)
fit7 <- lm(wage~poly(age,7,raw=TRUE),data=Wage)
fit8 <- lm(wage~poly(age,8,raw=TRUE),data=Wage)
fit9 <- lm(wage~poly(age,9,raw=TRUE),data=Wage)

anova(fit1,fit2,fit3,fit4,fit5,fit6,fit7,fit8,fit9)
```
`anova()` also shows that the minimul number of degree of freedom is 3.

Plot the fit with the determined number of $df$.
```{r}
agelims=range(Wage$age)
age.grid=seq(from=agelims[1],to=agelims[2])

fit<- glm(wage~poly(age,3,raw=T),data=Wage)
pred <- predict(fit, newdata=list(age=age.grid),se=TRUE)
se.bands <- cbind(pred$fit+2*pred$se  , pred$fit-2*pred$se )

plot(Wage$age,Wage$wage, main="Wage vs age with df=3", col="darkgray", pch=20, cex=0.7 )
lines(age.grid, pred$fit, lwd=2, col="blue")
matlines(age.grid, se.bands,col="blue", lty=2)    
```

### 6.b

```{r}
cutmax <- 10
all.cvs = rep(NA, cutmax)
for (i in 2:10) {
  Wage$age.cut = cut(Wage$age, i)
  lm.fit = glm(wage~age.cut, data=Wage)
  all.cvs[i] = cv.glm(Wage, lm.fit, K=10)$delta[2]
}
plot(2:10, all.cvs[-1], xlab="Number of cuts", ylab="CV error", type="l", pch=20, lwd=2)
```

Number of cuts is 8

```{r}
fit <- glm(wage~cut(age,8), data=Wage)
preds <- predict(fit, newdata=list(age=age.grid))
plot(wage~age, data=Wage, col="darkgrey", cex=0.8)
lines(age.grid, preds,  lwd=2,col="blue")
```

### Ex 7

```{r}
plot(wage~maritl, data=Wage)
plot(wage~jobclass, data=Wage)
```

From plots it appears a married couple makes more money on average than other groups. It
also appears that Informational jobs are higher-wage than Industrial jobs on
average.



#### gams
```{r}
require(gam)
gam1<- gam(wage~maritl+jobclass+s(age,df=4), data=Wage)
par( mfrow=c(1,3))
plot(gam1,se=TRUE)
```

#### Polynomial and Step functions
```{r 7.2}
fit = lm(wage~maritl, data=Wage)
deviance(fit)
fit = lm(wage~jobclass, data=Wage)
deviance(fit)
fit = lm(wage~maritl+jobclass, data=Wage)
deviance(fit)
```

### Ex 8

The flollow variables were identified as having non-linear relationship
```{r}
data(Auto)
par(mfrow=c(2,3))
plot(mpg~displacement, data=Auto)
plot(mpg~horsepower, data=Auto, cex=0.6)
plot(mpg~weight, data=Auto, cex=0.6)
plot(mpg~acceleration, data=Auto, cex=0.6)
plot(mpg~cylinders, data=Auto, cex=0.6)
```

Lets examine `cylinders`
```{r}
# polynomial regression
maxpd <-4
all.deltas = rep(NA, maxpd)
for (i in 1:maxpd) {
  glm.fit = glm(mpg~poly(cylinders, i), data=Auto)
  all.deltas[i] = cv.glm(Auto, glm.fit, K=10)$delta[2]
}
plot(1:maxpd, all.deltas, xlab="Degree", ylab="CV error", type="l", pch=20, lwd=2)

cylr <- range(Auto$cylinders)
cyl.grid <-seq(from=cylr[1], to=cylr[2])

cy.fit = glm(mpg~poly(cylinders, 3), data=Auto) # 
cy.preds <- predict(cy.fit, newdata=list(cylinders=cyl.grid),se=TRUE)
cy.se.bands=cbind(cy.preds$fit+2*cy.preds$se.fit ,cy.preds$fit-2*cy.preds$se.fit )

plot(mpg~cylinders,data=Auto)
lines(cyl.grid,cy.preds$fit,lwd=2,col="blue")
matlines(cyl.grid,cy.se.bands,lwd=1,col="blue",lty=3)
```





