---
title: "ch9_conceptual"
author: "S Zimine"

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

EX 1
=======
```{r}
x1 <- -10:10
x2 <- 1 + 3*x1
plot(x1,x2,type="l", col="red", ylim=range(x1), main="hyperplanes as lines in 2D")
abline(v=0,  lty="dotted")
abline(h=0,  lty="dotted")
text(c(5), c(-5), "greater than 0", col = "red")
text(c(-8), c(-5), "less than 0", col = "red")
x2_p <- 1 - 0.5*x1
lines(x1,x2_p, type="l", col="blue")
text(c(5), c(3), "greater than 0", col = "blue")
text(c(-8), c(3), "less than 0", col = "blue")
legend("topright", c("x2=1 + 3 * x1", "x2=1 - 0.5*x1"),col=c("red", "blue"), cex=1, lty=1 )
```


EX 2
======

$(1+X_1)^2 + (2-X_2)^2 =4$ is a circle with radius 2 and center (-1,2)

### a)
```{r}
radius = 2
plot(NA, NA, type = "n", xlim = c(-4, 2), ylim = c(-1, 5), asp = 1,
     xlab = "X1", ylab = "X2")
symbols(c(-1), c(2), circles = c(radius), add = TRUE,inches=FALSE)
abline(v=0,  lty="dotted")
abline(h=0,  lty="dotted")

```


###b)
```{r}
radius = 2
plot(NA, NA, type = "n", xlim = c(-4, 2), ylim = c(-1, 5), asp = 1,
     xlab = "X1", ylab = "X2")
symbols(c(-1), c(2), circles = c(radius), add = TRUE,inches=FALSE)
abline(v=0,  lty="dotted")
abline(h=0,  lty="dotted")
text(c(-1), c(2), "< 4")
text(c(-4), c(2), "> 4")
```

###c)
```{r}
radius = 2
plot(c(0, -1, 2, 3), c(0, 1, 2, 8), col = c("blue", "red", "blue", "blue"), 
    type = "p", asp = 1, xlab = "X1", ylab = "X2",pch=19)
symbols(c(-1), c(2), circles = c(radius), add = TRUE, inches = FALSE)
abline(v=0,  lty="dotted")
abline(h=0,  lty="dotted")
```

###d)
When we expand the equation in terms of quadratic terms the boudary is linear
$$(1+X_1)^2 + (2-X_2)^2 =4$$
$$1+ 2X_1 + X_1^2 + 4 - 4 X_2 + X_2^2 = 4$$
$$5+ 2X_1 + X_1^2 - 4 X_2 + X_2^2 = 4$$

EX 3
=====


###a)
```{r}
x1<-c(3,2,4,1,2,4,4)
x2<-c(4,2,4,4,1,3,1)
y<-c(rep(2,4),rep(4,3))

plot(x1,x2,col=y,pch=20, asp=1)
grid()
```


####b)
The optimal boundary appears to be $0.5 - X_1  + X_2 = 0$. 
```{r}
plot(x1,x2,col=y,pch=20, asp=1)
grid()
abline(a=-0.5,b=1, lty="dashed")
```

####c)
Classify to red if $0.5 - X_1  + X_2  < 0$ 

Classify to blue if $0.5 - X_1  + X_2  >0$ 

####d)
```{r}
plot(x1,x2,col=y,pch=20,asp=1)
grid()
abline(a=-0.5,b=1, lty="dashed")
abline(a=0,b=1, lty="dotted")
abline(a=-1,b=1, lty="dotted")
```

#####e)
finding support vectors
```{r}
which(x1%in%c(2,4))
which(x2%in%c(1,2,3,4))
```
We see that support points index ix `c(2,3,5,6)` Point 7 is not a support vector

```{r}
plot(x1,x2,col=y,pch=1,asp=1)
grid()
abline(a=-0.5,b=1, lty="dashed")
points(x1[c(2,3,5,6)],x2[c(2,3,5,6)], col=y[c(2,3,5,6)], pch=20)
```

####f)
A slight movement of observation 7 (4.1) would not move a separation plane as this points lies far away of the plane

A non-optimal hyperplane is not equally distant from support vectors
####g)
```{r}
plot(x1,x2,col=y,pch=1,asp=1)
grid()
abline(a=-0.9,b=1, lty="dashed")
```

###h
```{r}
plot(x1,x2,col=y,pch=1,asp=1)
grid()
points(c(4),c(2),col=2,pch=19) #additional point

```


